Tracing the Meta-Level: PyPy's Tracing JIT Compiler

### 2009-04-07 talk/icooolps2009/paper.tex
I got another round of review

\numberofauthors{3}
\author{
\alignauthor Carl Friedrich Bolz\\
       \affaddr{Heinrich-Heine-Universität Düsseldorf}\\
       \affaddr{STUPS Group}\\
       \affaddr{Deutschland}\\
       \email{cfbolz@gmx.de}
\alignauthor Antonio Cuni\\
       \affaddr{University of Genova}\\
       \affaddr{DISI}\\
       \affaddr{Italy}\\
       \email{cuni@disi.unige.it}
\alignauthor Armin Rigo\\
       \email{arigo@tunes.org}
\alignauthor Maciej Fijalkowski\\
       \email{fijal@merlinux.eu}
}
\maketitle

@@

\begin{abstract}
We present techniques for improving the results when a tracing JIT compiler is
applied to an interpreter. An unmodified tracing JIT performs not as well as one
would hope when the compiled program is itself a bytecode interpreter. We
examine why that is the case, and how matters can be improved by adding markers to
the interpreter, that help the tracing JIT to improve the results. We evaluate
the techniques by using them both on a small example as well as on a full Python
interpreter. This work has been done in the context of the PyPy project.

@@

The PyPy project is trying to find approaches to generally ease the
implementation of dynamic languages. It started as a Python implementation in
Python, but has now extended its goals to be generally useful for implementing
other dynamic languages as well. The general approach is to implement an
interpreter for the language in a subset of Python. This subset is chosen in
such a way that programs in it can be compiled into various target environments,
such as C/Posix, the CLI or the JVM. The PyPy project is described in more
details in Section \ref{sect:pypy}.

@@

The PyPy project\footnote{http://codespeak.net/pypy}
\cite{rigo_pypys_2006,carl_friedrich_bolz_to_2007} was started to implement a
new Python interpreter in Python but has now extended its goals to be an
environment where flexible implementation of dynamic languages can be written.
To implement a dynamic language with PyPy, an interpreter for that language has
to be written in RPython \cite{ancona_rpython:step_2007}. RPython (``Restricted
Python'') is a subset of Python
chosen in such a way that type inference can be performed on it. The language
interpreter can then be translated with the help of PyPy into various target
environments, such as C/Posix, the CLI and the JVM. This is done by a component
of PyPy called the \emph{translation toolchain}.

@@

A tracing JIT compiler finds the hot loops of the program it is compiling. In
our case, this program is the language interpreter. The most important hot interpreter loop
is the bytecode dispatch loop (for many simple
interpreters it is also the only hot loop).  Tracing one iteration of this
loop means that
the recorded trace corresponds to execution of one opcode. This means that the
assumption that the tracing JIT makes -- that several iterations of a hot loop
take the same or similar code paths -- is wrong in this case. It is very
unlikely that the same particular opcode is executed many times in a row.
\begin{figure}
\input{code/tlr-paper.py}
\vspace{-0.4cm}
\caption{A very simple bytecode interpreter with registers and an accumulator.}
\label{fig:tlr-basic}
\end{figure}
\begin{figure}
{\small
\begin{verbatim}
    MOV_A_R     0   # i = a
    MOV_A_R     1   # copy of 'a'

@@

Let's look at an example. Figure \ref{fig:tlr-basic} shows the code of a very
simple bytecode interpreter with 256 registers and an accumulator. The
\texttt{bytecode} argument is a string of bytes, all register and the
accumulator are integers. A program for this interpreter that computes
the square of the accumulator is shown in Figure \ref{fig:square}. If the
tracing interpreter traces the execution of the \texttt{DECR\_A} opcode (whose
integer value is 7), the trace would look as in Figure \ref{fig:trace-normal}.
Because of the guard on \texttt{opcode0}, the code compiled from this trace will
be useful only when executing a long series of \texttt{DECR\_A} opcodes. For all
the other operations the guard will fail, which will mean that performance is
probably not improved at all.

@@

To improve this situation, the tracing JIT could trace the execution of several
opcodes, thus effectively unrolling the bytecode dispatch loop. Ideally, the
bytecode dispatch loop should be unrolled exactly so much, that the unrolled version
corresponds to a \emph{user loop}. User loops
occur when the program counter of the \emph{language interpreter} has the
same value several times. This program counter is typically stored in one or several
variables in the language interpreter, for example the bytecode object of the
currently executed function of the user program and the position of the current
bytecode within that.  In the example above, the program counter is represented by 
the \texttt{bytecode} and \texttt{pc} variables.

@@

The language interpreter uses a similar techniques to detect \emph{hot user
loops}: the profiling is done at the backward branches of the user program,
using one counter per seen program counter of the language interpreter.

@@

Of course there is a lot of work still left to do. Various optimizations are not
quite finished. Both tracing and leaving machine code is very slow due to a
double interpretation overhead and we might need techniques for improving those.
Furthermore we need to apply the JIT to the various interpreters that are
written in RPython to evaluate how widely applicable the described techniques
are. Possible targets for such an evaluation would be the SPy-VM, a Smalltalk
implementation \cite{bolz_back_2008}, a Prolog interpreter or PyGirl, a Gameboy
emulator \cite{XXX}; but also less immediately obvious ones, like Python's
regular expression engine.

>>>>

\numberofauthors{3}
\author{
\alignauthor Carl Friedrich Bolz\\
       \affaddr{Heinrich-Heine-Universität Düsseldorf}\\
       \affaddr{STUPS Group}\\
       \affaddr{Germany}\\
       \email{cfbolz@gmx.de}
\alignauthor Antonio Cuni\\
       \affaddr{University of Genova}\\
       \affaddr{DISI}\\
       \affaddr{Italy}\\
       \email{cuni@disi.unige.it}
\alignauthor Armin Rigo\\
       \email{arigo@tunes.org}
\alignauthor Maciej Fijalkowski\\
       \email{fijal@merlinux.eu}
}
\maketitle

@@

\begin{abstract}
We present techniques for improving the results when a tracing JIT compiler is
applied to an interpreter.XXXAbrupt start - is the relevanz really immediately clear?XXX An unmodified tracing JIT performs not as well as one
would hope when the compiled program is itself a bytecode interpreter. We
examine the reasons for that, and how matters can be improved by adding markers to
the interpreter that help the tracing JIT to improve the results. We evaluate
the techniques by using them both on a small example as well as on a full Python
interpreter. This work has been done in the context of the PyPy project.

@@

The PyPy project is trying to find approaches to generally ease the
implementation of dynamic languages. It started as a Python implementation in
Python, but has now extended its goals to be generally useful for implementing
other dynamic languages as well. The general approach is to implement an
interpreter for the language in a subset of Python. This subset is chosen in
such a way that programs in it can be compiled into various target environments,
such as C/Posix, the CLI or the JVM. The PyPy project is described in more
detail in Section \ref{sect:pypy}.

@@

The PyPy project\footnote{http://codespeak.net/pypy}
\cite{rigo_pypys_2006,carl_friedrich_bolz_to_2007} is an
environment where flexible implementation of dynamic languages can be written.
To implement a dynamic language with PyPy, an interpreter for that language has
to be written in RPython \cite{ancona_rpython:step_2007}. RPython (``Restricted
Python'') is a subset of Python
chosen in such a way that type inference can be performed on it. The language
interpreter can then be translated with the help of PyPy into various target
environments, such as C/Posix, the CLI and the JVM. This is done by a component
of PyPy called the \emph{translation toolchain}.

@@

A tracing JIT compiler finds the hot loops of the program it is compiling. In
our case, this program is the language interpreter. The most important hot interpreter loop
is the bytecode dispatch loop (for many simple
interpreters it is also the only hot loop).  Tracing one iteration of this
loop means that
the recorded trace corresponds to execution of one opcode. This means that the
assumption made by the tracing JIT -- that several iterations of a hot loop
take the same or similar code paths -- is wrong in this case. It is very
unlikely that the same particular opcode is executed many times in a row.
\begin{figure}
\input{code/tlr-paper.py}
\vspace{-0.4cm}
\caption{A very simple bytecode interpreter with registers and an accumulator.}
\label{fig:tlr-basic}
\end{figure}
\begin{figure}
{\small
\begin{verbatim}
    MOV_A_R     0   # i = a
    MOV_A_R     1   # copy of 'a'

@@

An example is given in Figure \ref{fig:tlr-basic}. It shows the code of a very
simple bytecode interpreter with 256 registers and an accumulator. The
\texttt{bytecode} argument is a string of bytes, all register and the
accumulator are integers. A program for this interpreter that computes
the square of the accumulator is shown in Figure \ref{fig:square}. If the
tracing interpreter traces the execution of the \texttt{DECR\_A} opcode (whose
integer value is 7), the trace would look as in Figure \ref{fig:trace-normal}.
Because of the guard on \texttt{opcode0}, the code compiled from this trace will
be useful only when executing a long series of \texttt{DECR\_A} opcodes. For all
the other operations the guard will fail, which will mean that performance is
probably not improved at all.

@@

To improve this situation, the tracing JIT could trace the execution of several
opcodes, thus effectively unrolling the bytecode dispatch loop. Ideally, the
bytecode dispatch loop should be unrolled exactly so much that the unrolled version
corresponds to a \emph{user loop}. User loops
occur when the program counter of the \emph{language interpreter} has the
same value several times. This program counter is typically stored in one or several
variables in the language interpreter, for example the bytecode object of the
currently executed function of the user program and the position of the current
bytecode within that.  In the example above, the program counter is represented by 
the \texttt{bytecode} and \texttt{pc} variables.

@@

The language interpreter uses a similar technique to detect \emph{hot user
loops}: the profiling is done at the backward branches of the user program,
using one counter per seen program counter of the language interpreter.

@@

Of course there is a lot of work still left to do XXX fix this phrase. Various optimizations are not
quite finished. Both tracing and leaving machine code are very slow due to a
double interpretation overhead and we might need techniques for improving those.
Furthermore we need to apply the JIT to the various interpreters that are
written in RPython to evaluate how widely applicable the described techniques
are. Possible targets for such an evaluation would be the SPy-VM, a Smalltalk
implementation \cite{bolz_back_2008}, a Prolog interpreter or PyGirl, a Gameboy
emulator \cite{XXX}; but also not immediately obvious ones, like Python's
regular expression engine.
